
### Basic settings
```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(lattice)
library(ggplot2)
library(plyr)
```

## Loading and preprocessing the data

```{r}
# Read the Activity Data
unzip("activity.zip")
activity <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
activity$date <- as.Date(activity$date , format = "%Y-%m-%d")

#   Histogram of Total Number of Steps Taken Each Day -ignore the missing values in the dataset.
stepsperday<-ddply(activity, c("date"),summarise,
                   totalsteps=sum(steps,na.rm=TRUE)
)
ggplot(stepsperday, aes(date, totalsteps)) + geom_bar(stat = "identity",color="green", width = 0.8) + 
    labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")

```

## What is mean total number of steps taken per day?

```{r}
# Calculate the mean and median of the steps per day
stepsperdaydf <- aggregate(activity$steps, by=list(activity$date), sum)
names(stepsperdaydf)[1] <-"day"
names(stepsperdaydf)[2] <-"steps"
mean(stepsperdaydf$steps, na.rm = TRUE)
median(stepsperdaydf$steps, na.rm = TRUE )
```
## What is the average daily activity pattern?

```{r}
# 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
stepsperinterval<-ddply(activity, c("interval"),summarise,
                        meansteps = mean(steps,na.rm=TRUE)
)

plot(stepsperinterval$interval, stepsperinterval$meansteps, type="n", 
     main="Time Series Plot per 5-minute interval",
     xlab = "5-minute intervals",
     ylab = "Average number of steps taken") 
lines(stepsperinterval$interval, stepsperinterval$meansteps,type="l") 

```

```{r}
#  2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
stepsperinterval[stepsperinterval$meansteps == max(stepsperinterval$meansteps), ]
```
## Imputing missing values

```{r}
# 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
sum(!is.na(activity$steps))  # 2304/17568
```

```{r}
# 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
# 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
activitynew <- activity 
for (i in 1:nrow(activitynew)) {
    if (is.na(activitynew$steps[i])) {
        activitynew$steps[i] <- stepsperinterval[which(activitynew$interval[i] == stepsperinterval$interval), ]$meansteps
    }
}

head(activitynew)
sum(is.na(activitynew))

ggplot(activitynew, aes(date, steps)) + geom_bar(stat = "identity",width = 0.8) + 
     labs(title = "Total Number of Steps Taken Each Day (includes missing values)", x = "Date", y = "Total number of steps")
```

```{r}
# 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
stepsperdaydfnew <- aggregate(activitynew$steps, by=list(activitynew$date), sum)
names(stepsperdaydfnew)[1] <-"day"
names(stepsperdaydfnew)[2] <-"steps"
```
#   (NA's omitted)
The mean total number of steps taken per day is `r mean(stepsperdaydf$steps, na.rm = TRUE)`. 
The median number of steps taken per day is `r median(stepsperdaydf$steps, na.rm = TRUE )`.
#   (NA's imputed with mean values)
The mean total number of steps taken per day is `r mean(stepsperdaydfnew$steps, na.rm = TRUE)`. 
The median number of steps taken per day is `r median(stepsperdaydfnew$steps, na.rm = TRUE )`.
The Mean has remained the same but the Median has increased a little after NA replaced

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating 
#     whether a given date is a weekday or weekend day.
activitynew$weekday<-as.factor(ifelse(weekdays(activitynew$date)%in%c("Monday","Tuesday","Wednesday","Thursday","Friday"),"weekday","weekend"))
head(activitynew)

# 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
#     the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
stepsperinterval.weeksplit<-ddply(activitynew, c("interval","weekday"),summarise,
                                     meansteps = mean(steps,na.rm=TRUE)
)

xyplot(stepsperinterval.weeksplit$meansteps ~ stepsperinterval.weeksplit$interval | stepsperinterval.weeksplit$weekday, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```